---
title: "Lab 2: Probability Distributions"
subtitle: "Graded Out of 35 Points"
author: Jennifer Bradley
date: 02-11-25
output:
  pdf_document:
    latex_engine: xelatex
editor: visual
---

We will again be using our Palmer Penguins data today for some problems, so let's load that package in now:

```{r}
#| label: load-packagesa
#| include: true

library(palmerpenguins)
```

## 1. Working With Probability Distributions

### Problem 1.1 (5 Points)

Let $X$ be a discrete random variable whose PDF is described in the table given here:

| $X$    | 1   | 2   | 3   | 4   |
|--------|-----|-----|-----|-----|
| $f(X)$ | 3/8 | 1/8 | 1/8 | 3/8 |

: PDF of a discrete random variable, X

Do the following (show work, 1 point each):

A.  Calculate $E[X]$

    $\sum_{Sx} xp(x)$

    $(1\cdot\frac{3}{8}) + (2\cdot\frac{1}{8})+(3\cdot\frac{1}{8})+(4\cdot\frac{3}{8})$

    $$\boxed{2.5}$$

B.  Calculate $Var(X)$

    $\sum_{Sx} x^2p(x)-(\sum_{Sx} xp(x))^2$

    $((1^2\cdot\frac{3}{8}) + (22\cdot\frac{1}{8})+(32\cdot\frac{1}{8})+(42\cdot\frac{3}{8}))-2.52$

    $$
    \boxed{1.75}
    $$

C.  Plot $f(x)$. To make a barplot, use the `barplot` function.

    ```{r}
    fx <- c(3/8,1/8,1/8,3/8)
    X <- c(1,2,3,4)
    # realized I do not need this: prob1C.table <- data.frame(fx,X)

    barplot(fx, xlab="X", ylab="f(x)", names.arg = c("1","2","3","4"), ylim = c(0,1))
    ```

D.  Plot $F(x)$. To make a barplot, use the `barplot` function.

    ```{r}
    Fx <- c(3/8,((3/8)+(1/8)),((3/8)+(1/8)+(1/8)),((3/8)+(1/8)+(1/8)+(3/8)))
    # realized I do not need this: prob1D.table <- data.frame(Fx,X)

    barplot(Fx, xlab = "X", ylab = "f(x)", names.arg = c("1","2","3","4"))
    ```

E.  Calculate $P(X \geq 4)$

    $\text{4 is the only possible outcome $\le$ 4 in the set}$

    $P(4) = \frac{3}{8}$

    $$
    \boxed{\frac{3}{8}}
    $$

### Problem 1.2 (5 Points)

**Some background:** The exponential distribution is the continuous analog of the geometric distribution, which is discrete. Consider a geometric random variable $X$ with parameter prob=0.5 and an exponential random variable $Y$ with parameter rate=0.5. Note that R has a standard naming scheme for functions in which PDFs are named “d” + (distribution suffix) and CDFs are named “p” + (distribution suffix). The distribution suffix for the geometric distribution is `geom` (*e.g.*, `dbiom`), and for the exponential is `exp` (*e.g*. `dexp`). There is a trick in this question; think carefully!

A)  (2 Points) Plot the PDF for $X$ and $Y$ across a range from 0 to 10. One plot should be a barplot and the other a scatter plot with lines; part of the question is figuring out which variable matches which style of plot. Some tips:
    -   To generate a vector of evenly spaced values over a set interval, try the `seq` function.
    -   If if `v` is a vector with the values you want to evaluate between 0 and 10, then you can generate a vector with all the corresponding probabilities via `dgeom(x=v, prob=0.5)` and analogously for the exponential.
    -   To make a line plot, use the `plot` function with the option `type="l"`

```{r}
support0_10 <- c(seq(0,10))

XPDF <- function(x,p){
  ((1-p)^x)*p
}

barplot(XPDF(support0_10,0.5), ylab = "probability density", xlab = "x", main = "PDF for X")


```

```{r}
YPDF <- function(y,l){
  l*(exp(1)^(-l*y))
}

plot(YPDF(support0_10,0.5), type = "l", ylab = "probability density", xlab = "y", main = "PDF for Y")
```

A)  (2 Points) Calculate the probability that $X=5$ and the probability that $Y=5$

```{r}
XPDF(5,0.5)
```

::: callout-note
## Solution to 1.2B

P(X = 5) = 0.015625

P(Y = 5) = 0, because the probability that a random variable equals an exact number over a continuous interval is equal to 0
:::

A)  (1 Points) Calculate the probability that $5<X \leq 6$ and the probability that that $5 < Y \leq 6$. Hint: Use the CDF for this.

```{r}
pgeom(support0_10[7], prob=0.5) - pgeom(support0_10[6], prob=0.5)
pexp(support0_10[7], rate=0.5) - pexp(support0_10[6], rate=0.5)

```

::: callout-note
## Solution to 1.2C

\$ P(5\<X\leq6) = \boxed{0.0078125} \$

\$ P(5\<Y\leq6)=\boxed{0.03229793} \$
:::

## 2. Knowing Your Distributions

### Problem 2.1 (5 Points)

For each of the following situations, propose a probability distribution that would suit the situation and justify your selection (1 point each).

A.  We are interested in simulating the number of aphids per leaf on a tomato plant. For each leaf, there will be a discrete number of aphids, with no strict upper limit.

    A.  Poisson

        A.  Number of aphids –\> discrete

        B.  Fixed interval of space (per leaf) –\> Poisson

B.  We are interested in simulating the time it takes a vulture to locate it's next meal. Assume most vultures find a meal rather quickly, but some will take a very (arbitrarily) long time.

    A.  Exponential

        A.  Time –\> continuous

        B.  Amount of time it takes before a "success" (ie. meal located) –\> Exponential

C.  We have collected 100 moths from the environment that may belong to one of two different color morphs (light and dark). We are interested in simulating the number of light moths collected in this sample.

    A.  Binomial

        A.  Two discrete categories

        B.  Counting the number of one of two binary options –\> Binomial

D.  We are in the field, again sampling a species of moth that may belong to one of two color morphs (light or dark). We are interested in simulating whether or not an individual moth sampled will be from the dark color morph.

    A.  Bernoulli

        A.  Two discrete options

        B.  Counting one outcome out of two binary possibilities –\> Bernoulli

E.  We are interested in simulating the average number of leaves per tree in a forest across many different forests in the Northeast US.

    A.  Normal

        A.  Average number of leaves –\> continuous

        B.  Distribution of characteristic with additive effects –\> Normal

### Problem 2.2 (10 Points)

For each of the following, if necessary identify the probability distribution you are using and justify your choice, then answer the question showing your work.

A.  (2 Points) What is the probability that a predator will capture it's first prey item after six failed attempts given that it's probability of capturing prey per attempt is 0.3?

    \$\text{PMF of a geometric distribution =} (1-p)\^x\cdot p \$

    \$ P(\text{capture on 7th attempt) =} (1-0.3)\^{6}\cdot0.3 \$

    \$ = \boxed{0.0352947} \$

B.  (2 Points) If new members of a species immigrate into a defined patch of habitat at a constant rate of 0.5 per day, what is the expected wait time until the next new arrival?

    \$ \text{Expected Value x of an exponential distribution}= \frac{1}{lambda}}\$

    \$ = \boxed{2} \$

C.  (2 Points) For the example above (B), what is the expected number of arrivals over a one-week period?

    Poisson

    \$0.5\\cdot7\\text{ days}=\\boxed{3.5}\$

D.  (3 Points) We randomly selected 100 White Ash trees from a forest in upstate NY. For the area we are interested in, we know that from previous surveys about 30% of trees are infested by Emerald Ash Borers (assume a 0.3 probability that any given tree is infested). Use R to plot the pdf of the number of infested trees in your sample. Discuss (briefly!) the limitations of our approach for modeling this probability - how might the assumptions of our model be violated?

    This approach assumes that cases of infection are independent. However, in a real-world case it is possible that trees can spread disease to one another (ie. the infection of one tree leads directly to the infection of a tree in its proximity).

    ```{r}
    whiteashtrees <- seq(0,100, by = 1)
    plot(dbinom(whiteashtrees, 100, 0.3), ylim = c(0,.3), ylab = "Probability of Infection", xlab = "Sample", main = "PDF of 100 White Ash Trees with Parameter p = 0.3")
    ```

E.  (1 Point) We are studying a forest where tree heights are normally distributed with a mean height of 10 meters and a variance of 4 meters. What percent of trees do we expect to fall within a height range of 6-14 meters?

    ::: callout-note
    ## 2.2E

    95% of samples in a normal distribution are within +- 2 standard deviations from the mean. Therefore, 95% of trees would be between 6 and 14 meters (10 +- 4 meters).
    :::

## 3. Checking for Normality with Quantile-Quantile Plots

### Problem 3.1 (5 Points)

::: callout-note
## 3.1

The Q-Q plot does not fit the trendline very well (especially at small and large values), which does not support that the data is exactly normally distributed.
:::

```{r}
lnormal_qqplot <- function(data, x){
  ldata <- log(data)
  qqnorm(ldata, main = x)
  qqline(ldata)
}

lnormal_qqplot(data = penguins$body_mass_g, "Q-Q Body Mass")
```

### Problem 3.2 (5 Points)

```{r}
a_penguins <- subset(penguins, penguins$species=="Adelie")
cs_penguins <- subset(penguins, penguins$species=="Chinstrap")
g_penguins <- subset(penguins, penguins$species=="Gentoo")

par(mfrow=c(1,3))
lnormal_qqplot(a_penguins$body_mass_g, "Adelie Penguin Body Mass")
lnormal_qqplot(cs_penguins$body_mass_g, "Chinstrap Penguin Body Mass")
lnormal_qqplot(g_penguins$body_mass_g, "Gentu Penguin Body Mass")


female_penguins <- subset(penguins, sex == "female")
male_penguins <- subset(penguins, sex == "male")

par(mfrow=c(1,2))
lnormal_qqplot(female_penguins$body_mass_g, "Female Penguins Body Mass")
lnormal_qqplot(male_penguins$body_mass_g, "Male Penguins Body Mass")

female_a_penguins <- subset(a_penguins, sex == "female")
male_a_penguins <- subset(a_penguins, sex == "male")

female_cs_penguins <- subset(cs_penguins, sex == "female")
male_cs_penguins <- subset(cs_penguins, sex == "male")

female_g_penguins <- subset(g_penguins, sex == "female")
male_g_penguins <- subset(g_penguins, sex == "male")

par(mfrow=c(2,3))
lnormal_qqplot(female_a_penguins$body_mass_g,"Female Adelie Penguins Body Mass")
lnormal_qqplot(male_a_penguins$body_mass_g,"Male Adelie Penguins Body Mass")

lnormal_qqplot(female_cs_penguins$body_mass_g, "Female Chinstrap Penguins Body Mass")
lnormal_qqplot(male_cs_penguins$body_mass_g,"Male Chinstrap Penguins Body Mass")

lnormal_qqplot(female_g_penguins$body_mass_g, "Female Gentu Penguins Body Mass")
lnormal_qqplot(male_g_penguins$body_mass_g, "Male Gentu Penguins Body Mass")

```

## 5. List of All Problems (Above, 35 Points)

-   1.1(5 Points)

-   1.2 (5 Points)

-   2.1 (5 Points)

-   2.2 (10 Points)

-   3.1 (5 Points)

-   3.2 (5 Points)

## 6. Acknowledgements

Problem 1.2 adapted from problem sets shared by Dr. Philip L.F. Johnson. Problem 1.1 adapted from problem sets shared by Dr. Heather Lynch.
